---
title: "Spatial Autocorrelation"
author: "Jakob Grøhn Damgaard"
date: "3/28/2021"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r loading packages, include=FALSE}
# Loading libraries
library(raster)
library(rgeos)
library(sf)
library(tidyverse)
library(htmltools)
library(googlesheets4)
library(mapview)
library(tmap)
library(spdep)
library(cartogram)
```

# Introduction
The purpose of this script is to analyse and check for spatial autocorrelation in the distribution of unmarried men in Denmark. The spatial analysis is performed at a municipality level using Moran's I test.
<br>

## Loading data
Loading in both the map containing municipality multipolygons and civilstatus data for Danish citizens.
```{r loading data, warning=FALSE}
# Load data fir the Danish municipalities
mun_sp <- readRDS("data/gadm36_DNK_2_sp.rds")
# Transform to an sf object
mun_sf <- st_as_sf(mun_sp)
# Convert CRS
mun <- st_transform(mun_sf, crs = 32632)
# Plotting the municipalities to interactive map using mapview()
#mapview(mun)

# Changing some of the names in order to merge data sets later
sort(mun$NAME_2)
which(grepl("Å",mun$NAME_2))
which(grepl("Taastrup",mun$NAME_2))
which(grepl("Vest",mun$NAME_2))
# Renaming the relevant names
mun$NAME_2[31] <- "Aarhus"
mun$NAME_2[21] <- "Høje-Taastrup"
mun$NAME_2[60] <- "Vesthimmerlands"

#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# Loading data on the civilstatus of Danish citizens
civilstatus_DK <- read_csv("data/civilstatus_DK.csv")
```

## Data pre-processing
Pre-processing the civilstatus data to make it fit for analysis.
```{r pressure, echo=FALSE}
# Selecting only male 2020 data and removing entrances containing data on a regional level or rows with NA values
male_civilstatus_DK_2020 <- civilstatus_DK %>% select("Status", "Sex", "Region", "Y2020K1") %>% filter(Sex == "Mænd") %>% filter(!grepl("Region",Region)) %>% drop_na(Region)

# Saving the total population of each municipality in the data set in 2020 in order to calculate the percentage of unmarried
male_pop2020 <- male_civilstatus_DK_2020 %>% 
  group_by(Region) %>% 
  summarize(pop2020 = sum(Y2020K1))

# Merge data sets, keep only unmarried and create new column containing unmarried in percentage of population
unmarried_men_pr_region <- mun %>% 
  select(NAME_2) %>% 
  merge(male_pop2020, by.x = "NAME_2",by.y ="Region") %>% 
  merge(male_civilstatus_DK_2020, by.x = "NAME_2",by.y ="Region") %>% 
  group_by(NAME_2, Sex) %>% 
  filter(Status == "Ugift") %>%
  mutate(pct_unmarried = Y2020K1/pop2020*100)
```

<br> The pre-processing is done. Before proceeding, we should quickly asses whether the data looks alright.
<br>
## Investigating the data
```{r plotting}
# Investigating the
unmarried_men_pr_region[1:6,]

# Plotting a map where the unicipalities are colored according to percentage of unmarried men
 unmarried_men_pr_region %>% 
  tm_shape() + 
  tm_polygons("pct_unmarried",
              title= "Percentage of unmarried men in 2020")
```

<br> Everything appears to look fine. Aarhus Kommune seems to have an extraordinary high percentage of unmarried men. As a large part of the population consists of students, this is in accordance with our expectations. Now we can proceed to perform the statistical correlation analysis using Moran's I test.

## Spatial Autocorrelation analysis 
```{r}
# Simplifying the multipolygon boundaries to minimise compute intensity
mun_simple <- st_cast(st_simplify(mun, dTolerance = 250), to = "MULTIPOLYGON")

# Create neighbour list following queen adjacency
nb <- poly2nb(mun_simple$geometry)

# Get center points of each municipality
mun_centers <- st_coordinates(st_centroid(mun_simple$geometry))

# Plot map with the intermunicipality connections
plot(mun_simple$geometry); plot(nb, mun_centers, col = "red",add = TRUE)

# Perform analytical Moran I test on percentage of unmarried men in 2020 across the 99 Danish municipalities
moran.test(unmarried_men_pr_region$pct_unmarried, 
           nb2listw(nb, style = "W",zero.policy=TRUE),
           zero.policy=TRUE)

# Perform Moran I test on percentage of unmarried men in 2020 across the 99 Danish municipalities using MC simulations
moran.mc(unmarried_men_pr_region$pct_unmarried, 
         nb2listw(nb, zero.policy=TRUE),
         zero.policy=TRUE, 
         nsim = 999)
```

# Conclusion
Both the analytical approach and the simulation based approach for analysing the spatial correlation yielded  z-scores (slightly negative ~ -0.006) close to 0 and both scores were not significant (p>0.05). This means that we cannot reject the null hypothesis and, hence, suggests that the distribution of unmarried men in 2020 is not spatially correlated at a municipality level. Thus unmarried men are spatially distributed randomly across Denmark and the are no inter-dependencies across municipality borders. <br>
It would be interesting to run a similar analysis at a more fine-grained level of geographical tessellation (e.g. zip codes) or with different definitions of neighbours.